---
title: "Initial analysis of LD loss simulations"
author: "Gibran Hemani"
date: "`r Sys.Date()`"
format:
  html:
    embed-resources: true
---

LD loss analysis asks

Simulations:

1. For a causal variant being discovered in European datasets, what are the possible top GWAS hits
2. What is the probability of each SNP being top hit in the region
3. What is the average expected LD between causal variant and European-discovered top hit in non-Europeans
4. What is the difference in allele frequencies between Eur and non-Eur for the Eur top hit

Analysis:

5. Are the known GWAS hits giving a different cross-population result from other LD/MAF matched SNPs
6. What is expected influence on MR
  - Using European bgx against non-European bgy
  - Using re-estimated bgx against bgy in non-Europeans
  - Consider bias due to LD, weak instrument bias, power


```{r}
library(here)
# renv::activate(here(".."))
library(dplyr)
library(ggplot2)
library(lmerTest)
library(lme4)
library(simulateGP)
```

```{r}
load(here("../output_set1/results.rdata"))
# remove duplicate SNPs

res <- res[!duplicated(paste(res$snp, res$pop, res$pval)),]
res$rsidcode <- paste(res$snp, res$pval)
all(table(paste(res$snp, res$pval)) == 5)
dim(res)
```

LD loss

```{r}
res %>% 
  mutate(pvalbin = cut(-log10(pval), 5), nbin=cut(n, 5)) %>%
  ggplot(., aes(y=r2, x=pop)) +
  geom_boxplot() +
  facet_grid(nbin ~ pvalbin)
```

The sample size doesn't have much impact on the decay of LD. But the discovery p-value has a big impact it seems - at least on an absolute scale

```{r}
res %>% 
  mutate(pvalbin = cut(-log10(pval), 5), nbin=cut(n, 5)) %>%
  ggplot(., aes(y=r2, x=pvalbin)) +
  geom_boxplot(aes(fill=pop)) +
  scale_fill_brewer(type="qual")
```

What about relative LD loss compared to EUR?
```{r}
res <- res %>%
  group_by(snp, pval) %>%
  mutate(r2_rel = r2/r2[pop=="EUR"])

res %>%
  group_by(pop) %>%
  summarise(r2_rel=mean(r2_rel))
```


```{r}
res$pvalbin <- cut(-log10(res$pval), 10)
res %>% 
  ggplot(., aes(y=r2_rel, x=pvalbin)) +
    geom_boxplot(aes(fill=pop)) +
    scale_fill_brewer(type="qual")
```

Plot relative LD decay by P-value for each population

```{r}
res %>% 
  group_by(pop, pvalbin) %>%
  summarise(r2_rel=mean(r2_rel)) %>%
  ggplot(., aes(x=pvalbin, y=r2_rel, group=pop)) +
  geom_line(aes(colour=pop))
```

It seems like the reduction in r2 is related to p-value even when looking at relative reduction. So what is important - relative LD decay or absolute LD decay? 

Small sample size - Europeans have a chance of low LD, and AFR variant could be very low.
Large sample size - Europeans more likely high LD, makes AFR LD closer, but the relative decay is still a problem.
So if instrumenting using EUR exp and testing AFR outcome, as sample size increases the absolute difference in LD is going to be better.

Let's look at allele frequency differences now

```{r}
res$maf <- res$af
res$maf[res$maf > 0.5] <- 1 - res$maf[res$maf > 0.5]
ggplot(res, aes(x=maf)) +
  geom_histogram() +
  facet_wrap(~ pop)
```

European shows ascertainment for common variants. The corresponding allele frequency in AMR and SAS seems to match somewhat (surprising) but the AFR and EAS dist is much more uniform - many discovered common EUR variants will be rare in EAS and AFR - reduces power, increase weak instrument bias

```{r}
res %>%
  group_by(pop) %>%
  summarise(varsnp=mean(2*maf*(1-maf)))
```

How does allele frequency change with pval

```{r}
res %>% 
  group_by(pop, pvalbin) %>%
  summarise(maf=mean(maf)) %>%
  ggplot(., aes(x=pvalbin, y=2*maf*(1-maf), group=pop)) +
  geom_line(aes(colour=pop))
res$pvalbin <- cut(-log10(res$pval), 10)
res %>% 
  ggplot(., aes(y=2*maf*(1-maf), x=pvalbin)) +
    geom_boxplot(aes(fill=pop)) +
    scale_fill_brewer(type="qual")
```

No obvious relationship with MAF and pval

```{r}
summary(lm(maf ~ I(-log10(pval)) * pop, res))
```

```{r}
res %>% 
  ggplot(., aes(y=2*maf*(1-maf), x=pop)) +
    geom_boxplot(aes(fill=pop)) +
    scale_fill_brewer(type="qual")
```

```{r}
left_join(subset(res, pop != "EUR", select=c(rsidcode, maf, pop)), subset(res, pop == "EUR", select=c(rsidcode, maf)), by="rsidcode") %>%
  mutate(gvar.x=2*maf.x*(1-maf.x), gvar.y=2*maf.y*(1-maf.y)) %>%
  ggplot(., aes(x=gvar.y, y=gvar.x)) +
  geom_point() + 
  geom_smooth() +
  facet_wrap(~ pop) +
  labs(x="European var(SNP)", y="non-European var(SNP)")
```

```{r}
left_join(subset(res, pop != "EUR", select=c(rsidcode, maf, pop)), subset(res, pop == "EUR", select=c(rsidcode, maf)), by="rsidcode") %>%
  mutate(gvar.x=2*maf.x*(1-maf.x), gvar.y=2*maf.y*(1-maf.y)) %>%
  group_by(pop) %>%
  do({
    summary(lm(gvar.x ~ gvar.y, .))$coef %>% as_tibble() %>% slice_tail(n=2)
  })
```

Generally the MAF in non-Europeans is lower than discovery variants in Europeans. There is a relationship of higher MAF with larger p-value as expected.

Are the non-GWAS hit SNPs behaving differently in some way? 

```{r}
temp <- snplist %>%
  ungroup %>%
  select(rsid, p, target) %>%
  filter(!duplicated(paste(rsid, p))) %>%
  mutate(
    type = case_when(is.na(target) ~ "gwas", TRUE ~ "control")
  )
temp
# For simplicity just remove any control SNPs that are also gwas hits
temp2 <- subset(temp, type=="control")
rem <- temp2 %>% subset(rsid %in% subset(temp, type=="gwas")$rsid) %>% {.$rsid}
temp <- subset(temp, ! (rsid %in% rem & type=="control"))
temp
```

For every gwashit get the r2 and the average r2 for the control SNPs

```{r}
res$rsidcode <- paste(res$snp, res$pval)
temp$rsidcode <- paste(temp$rsid, temp$p)
x <- subset(temp, target == temp$rsid[1])

gwas_control <- temp %>% 
  filter(type == "control") %>% 
  group_by(target) %>%
  do({
    x <- .
    controlid <- x$rsidcode
    target <- paste(x$target[1], x$p[1])
    restemp <- subset(res, rsidcode %in% controlid) %>% 
      group_by(pop) %>%
      summarise(r2 = mean(r2), maf=mean(maf))
    tartemp <- subset(res, rsidcode == target)
    tibble(
      target = target,
      pop = c(tartemp$pop, restemp$pop),
      r2 = c(tartemp$r2, restemp$r2),
      maf = c(tartemp$maf, restemp$maf),
      what=rep(c("gwas", "control"), each=5)
    )
  })
gwas_control
```

```{r}
gwas_control %>%
  group_by(target, pop) %>%
  summarise(
    r2 = r2[what=="gwas"] - r2[what=="control"],
    maf = maf[what=="gwas"] - maf[what=="control"]
  ) %>%
  group_by(pop) %>%
  summarise(r2_mean=mean(r2), r2_se=sd(r2)/sqrt(n()), r2_p=pnorm(abs(r2_mean)/r2_se, low=F), maf_mean=mean(maf), maf_se=sd(maf)/sqrt(n()), maf_p=pnorm(abs(maf_mean)/maf_se, low=F))
```

The GWAS hits seem to have smaller LD decay than control SNPs in Africans. 

```{r}
gwas_control %>%
  group_by(target, pop) %>%
  summarise(
    r2 = r2[what=="gwas"] - r2[what=="control"],
    maf = maf[what=="gwas"] - maf[what=="control"]
  ) %>%
  group_by(pop) %>%
  left_join(., temp, by=c("target"="rsidcode")) %>%
  ggplot(., aes(x=-log10(p), y=r2)) +
#    geom_point(aes(colour=pop)) +
    geom_smooth(aes(colour=pop)) +
    scale_color_brewer(type="qual")
```

Insufficient power to notice a relationship with discovery p-value

```{r}
gwas_control %>%
  ggplot(., aes(x=pop, y=r2)) +
  geom_boxplot(aes(fill=what))
```

Visualisation that for each population, the LD decay for control SNPs seems to be larger than for known GWAS hits

```{r}
lmerTest::lmer(r2 ~ what + (1|pop), gwas_control) %>% summary()
#lme4::lmer(r2 ~ what + (1|pop), gwas_control) %>% summary()

```

## Influence on MR


### Genetic interactions

The ways in which change in LD and MAF can impact MR will differ based on how the MR is being performed. 

$$
\beta_{gy}^{AFR} / \beta_{gx}^{EUR} = \beta_{IV}R^2
$$
Core assumption for 2-sample MR is that causal structure in sample 1 is the same as in sample 2. However if the causal structure really is the same then LD differences could introduce biased effects if we don't have causal variants.

Difference in MAF could have an impact on genetic interactions, which in turn lead to differences in effects across populations.

If the effect of x on y is estimated using $\beta_{IV} = \beta_{gy}/\beta_{gx}$, then genetic or environmental interactions can modulate $\hat{\beta}_{gx}$ estimates as the allele frequencies change. For example

$$
x = a + \beta_{gx} G + \beta_2 G_2 + \beta_{int} GG_2 + e
$$

What is the expected association of x on G?

$$
\begin{aligned}
\hat{\beta}_{gx} &= cov(G, X)/Var(G) \\
&= cov(G, \beta_{gx}G + \beta_{int}GG_2)/Var(G) \\
&= \frac{\beta_{gx}Var(G) + 2p_2\beta_{int}Var(G)}{Var(G)} \\
&= \beta_{gx} + 2p_2\beta_{int}
\end{aligned}
$$

```{r}
sim_system <- function(n, af1, rho1, af2, rho2, b1, b2, bint)
{
  args <- environment() %>% as.list() %>% as_tibble()
  g1 <- simulateGP::correlated_binomial(n, af1, af1, rho1)
  g2 <- simulateGP::correlated_binomial(n, af2, af2, rho2)
  x <- g1[,1] * b1 + g2[,1] * b2 + g1[,1]*g2[,1]*bint
  bind_rows(
    tibble(
      coef=paste("SNP", c(1,2)),
      est=c(lm(x ~ g1[,1])$coef[2], lm(x ~ g2[,1])$coef[2]),
      ld="causal"
    ),
    tibble(
      coef=paste("SNP", c(1,2)),
      est=c(lm(x ~ g1[,2])$coef[2], lm(x ~ g2[,2])$coef[2]),
      ld="non-causal"
    )
  ) %>% bind_cols(., args) %>% return()
}

param <- expand.grid(af1=seq(0.1,0.9,by=0.1), af2=seq(0.1,0.9,by=0.1), n=1000, b1=1, b2=1, bint=c(0,1,2), rho1=0.8, rho2=0.8)
out <- lapply(1:nrow(param), function(i) do.call(sim_system, param[i,])) %>% bind_rows()
out %>% filter(bint == 2) %>%
  ggplot(., aes(x=af1, y=est, groups=as.factor(af2))) +
    geom_point(aes(colour=af2)) +
    geom_line(aes(colour=af2)) +
    facet_grid(ld ~ coef)
```

Effect of SNP1 depends on allele frequency of SNP2 and vice versa. When there is also imperfect LD this leads to the interaction effect being attenuated

Expected effect estimate of b_gx when it interacts with G2

$$
\begin{aligned}
\beta_{gx} &= cov(G,X)/var(G) \\
&= cov(G,\beta_{gx}G+\beta_{int}GG_2) /var(G) \\
&= \beta_{gx} + 2p_2\beta_{int}
\end{aligned}
$$

Check

```{r}
# cov(g1g2b1, g1)
g1 <- rbinom(10000, 2, 0.3)
g2 <- rbinom(10000, 2, 0.9)
b1 <- 0.5
cov(g1, g1*g2*b1)
b1*cov(g1*g2,g1)
b1*mean(g2)*var(g1)
x <- g1 + g2 + g1*g2*0.3
summary(lm(x ~ g1))
0.3*mean(g2)+1
```

Now what if we don't have the causal variant, tagging variant $G^*$ is correlated with causal variant by $\rho$

$$
\begin{aligned}
\beta_{G^*X} &= cov(G^*,X) / var(G^*)\\
&= cov(G^*, \beta_{GX}G+\beta_{int}GG_2)  / var(G^*)\\
&= [\beta_{GX}cov(G^*G) + 2p_2\beta_{int}cov(G^*G) ] / var(G^*) \\
&= [\rho var(G^*)\beta_{GX} + 2\rho p_2 var(G^*)\beta_{int}]  / var(G^*) \\
&= \rho\beta_{GX} + 2p_2 \rho \beta_{int}
\end{aligned}
$$

Check it here

```{r}
n <- 10000
p1 <- 0.1
p2 <- 0.1
rho <- sqrt(0.2)
g1 <- correlated_binomial(n, p1, p1, rho)
g2 <- correlated_binomial(n, p2, p2, rho)
x <- g1[,1] + g2[,1] + g1[,1]*g2[,1]

cov(g1[,1], x)
var(g1[,1]) + var(g1[,1])*mean(g2[,1])

cov(g1[,2], x)
cov(g1[,2], g1[,1] + g1[,1]*g2[,1])
cov(g1[,2], g1[,1]) + cov(g1[,2], g1[,1]*g2[,1])
cov(g1[,2], g1[,1]) + mean(g2[,1])*cov(g1[,2], g1[,1])
rho * 2*p1*(1-p1) + 2*p2*2*p1*(1-p1)*rho

lm(x ~ g1[,2])$coef[2]
rho + 2*p2*rho

```


So the difference in $\beta_{gx}$ across populations A and B will depend on a) the difference in frequencies across populations and b) the magnitude of $\beta_{int}$

$$
\hat{\beta}_{gx}^A - \hat{\beta}_{gx}^B = 2\beta_{int}(p_2^A-p_2^B)
$$

Note that the LD of tagging variants could modify the $\beta_{gx}$ difference

$$
\begin{aligned}

\hat{\beta}_{gx}^A - \hat{\beta}_{gx}^B &= \rho^A\beta_{GX} + 2p^A_2 \rho^A \beta_{int} - \rho^B\beta_{GX} - 2p^B_2 \rho^B \beta_{int} \\
&= \beta_{GX}(\rho^A - \rho^B) + 2\beta_{int}(\rho^A p_2^A - \rho^B p_2^B)
\end{aligned}
$$

Check this

```{r}
param <- expand.grid(n=10000, pA=seq(0.1,0.9,by=0.1), pB=seq(0.1,0.9,by=0.1), rho=c(0.999, 0.75, 0.5))
sim <- function(n, pA, pB, rho)
{
  args <- environment() %>% as.list() %>% as_tibble()
  gA1 <- correlated_binomial(n, 0.5, 0.5, rho)
  gA2 <- correlated_binomial(n, pA, pA, rho)
  gB1 <- correlated_binomial(n, 0.5, 0.5, rho)
  gB2 <- correlated_binomial(n, pB, pB, rho)
  xA <- gA1[,1] + gA2[,1] + gA1[,1] * gA2[,1]
  xB <- gB1[,1] + gB2[,1] + gB1[,1] * gB2[,1]
  
  yA <- xA + rnorm(n)
  yB <- xB + rnorm(n)
  
  # In pop1 use the causal variant, in pop2 the correlated
  out <- list()
  out$bgxA <- lm(xA ~ gA1[,1])$coef[2]
  out$bgyA <- lm(yA ~ gA1[,1])$coef[2]
  
  out$bgxB <- lm(xB ~ gB1[,2])$coef[2]
  out$bgyB <- lm(yB ~ gB1[,2])$coef[2]
  out$bgyB_d <- lm(yB ~ gB1[,1])$coef[2]
  
  out$bivA <-  out$bgyA / out$bgxA
  out$bivB <-  out$bgyB / out$bgxB
  out$bivB_d <-  out$bgyB_d / out$bgxB
  return(out %>% as_tibble() %>% bind_cols(args, .))
}
out2 <- lapply(1:nrow(param), function(i) do.call(sim, param[i,])) %>% bind_rows()
plot(bivA ~ bivB, out2)

out2 %>%
  mutate(
    bgxa_b = bgxA - bgxB,
    exp_bgxa_b = rho + 2*(pA-rho*pB)
  ) %>%
  filter(rho == 0.5) %>%
  {plot(exp_bgxa_b ~ bgxa_b, .)}
```

Note that the LD of tagging variants shouldn't have an impact here for the MR estimate - it will just cancel out

```{r}
n <- 100000
g1 <- correlated_binomial(n, 0.8,0.8, rho=0.5)
g2 <- correlated_binomial(n, 0.3,0.3, rho=0.5)
x <- g1[,1] + g2[,1] + g1[,1]*g2[,1]
y <- x + rnorm(n)

library(systemfit)
# Using causal variant as instrument
systemfit(y ~ x, method="2SLS", inst=~g1[,1])
# Using correlated variant as instrument
systemfit(y ~ x, method="2SLS", inst=~g1[,2])
```

From a cross ancestry perspective what this means is that if you use European instruments then they will have a different allele frequency and R2 in the non-European sample.

Scenario 1 - $\beta_{gy}^{AFR} / \beta_{gx}^{EUR}$

- LD differences 
- Allele frequency differences at interacting loci
- Weak instrument bias due to LD and allele frequency differences

$$
\begin{aligned}
\beta_{IV}^A &= \beta_{gy} / \beta_{gx}^A \\
&= \frac{\beta_{gy}}{\beta_{gx}\rho^2 + 2\beta_{int}\rho^4(p_2^A-p_2^B)} \\
\end{aligned}
$$

$$
\begin{aligned}
\hat{\beta}_{IV*} &= \frac{\beta_{gy}^{AFR}}{\beta_{gx}^{EUR}} \\
&= \frac{\beta_{IV}(\rho\beta_{GX} + 2p_2 \rho \beta_{int})}{\beta_{gx}^{EUR}}
\end{aligned}
$$

Scenario 2 - $\beta_{gy}^{AFR} / \beta_{gx}^{AFR}$ using g identified in Europeans

- Weak instrument bias due to LD and allele frequency differences

Scenario 3 - $\beta_{gy}^{AFR} / \beta_{gx}^{AFR}$ after multi-ancestry variant selection


### Weak instrument bias

Weak instrument bias is approximately $1/F$ where the F statistic is

$$
\begin{aligned}
F &= \frac{R_{gx}^2(n-2)}{(1-R_{gx}^2)} \\
&= \frac{2p(1-p)\beta_{gx}^2 \rho^2/Var(y)(n-2)}{1-2p(1-p)\beta_{gx}^2\rho^2/Var(y)}
\end{aligned}
$$

So an F statistic of 10 would lead to a weak instrument bias of 10%. When going from one population to another the instrument strength will change due to 

- reduced LD 
- lower allele frequency
- likely lower sample size
- no winner's curse

So the difference in F statistics will be

$$
\begin{aligned}
F_A - F_B &= \frac{2p_A(1-p_A)\beta_{gx}^2 \rho_A^2/Var(y_A)(n-2)}{1-2p_A(1-p_A)\beta_{gx}^2\rho_A^2/Var(y_A)} - \frac{2p_B(1-p_B)\beta_{gx}^2 \rho_B^2/Var(y_B)(n-2)}{1-2p_B(1-p_B)\beta_{gx}^2\rho_B^2/Var(y_B)}\\
\end{aligned}
$$

or the ratio of F statistics

$$
\begin{aligned}
\frac{F_A}{F_B} &= \frac{\frac{2p_A(1-p_A)\beta_{gx}^2 \rho_A^2/Var(y_A)(n_A-2)}{1-2p_A(1-p_A)\beta_{gx}^2\rho_A^2/Var(y_A)}} {\frac{2p_B(1-p_B)\beta_{gx}^2 \rho_B^2/Var(y_B)(n_B-2)}{1-2p_B(1-p_B)\beta_{gx}^2\rho_B^2/Var(y_B)}}\\
\end{aligned}
$$

For a set of European discovery hits, get a distribution of change in R2 and change in allele frequency per population. 

```{r}
res_eur <- subset(res, pop=="EUR") %>%
  mutate(
    rsq = TwoSampleMR::get_r_from_pn(pval, n)^2,
    beta = sqrt(rsq / (2*maf*(1-maf))/r2),
    Fstat = rsq * (n-2)/(1-rsq)
  )
res_eur %>%
  plot(Fstat ~ I(-log10(pval)), .)

res2 <- left_join(res, dplyr::select(ungroup(res_eur), rsidcode, beta), by="rsidcode") %>%
  mutate(
    rsq = beta^2 * r2 * 2 * maf * (1-maf)
  )
glimpse(res2)
```

Analysis

```{r}
res_eur <- subset(res2, pop=="EUR")
ratios <- lapply(c("EAS", "AFR", "SAS", "AMR"), function(p)
  {
    res_pop <- subset(res2, pop == p) %>% left_join(., res_eur, by="rsidcode") %>%
      mutate(
        rsq_ratio = rsq.x/rsq.y,
        Fstat_ratio = (rsq.x/(1-rsq.x))/(rsq.y/(1-rsq.y))
      ) %>%
      dplyr::select(rsidcode, n=n.x, rsq_ratio, Fstat_ratio, rsq_eur=rsq.y, rsq_noneur=rsq.x) %>%
      mutate(pop=p)
}) %>% bind_rows()
ratios
```

```{r}
summary(ratios$Fstat_ratio)
```

```{r}
subset(ratios, Fstat_ratio > 20)
```

## Difference in MAF and LD can have an impact on power

For equal sample size. How is power determined? Relates to R2 and sample size... Expected SE in exposure and outcome


```{r}
ratios %>%
  filter(rsq_ratio > quantile(rsq_ratio, 0.05, na.rm=T) & rsq_ratio < quantile(rsq_ratio, 0.95, na.rm=T)) %>%
  ggplot(., aes(y=rsq_ratio, x=pop)) +
    geom_violin() +
    scale_y_log10()
```



```{r}
str(out2)
```

$$
\beta_{IV} = \beta_{gy}/\beta_{gx}
$$

So 

```{r}
param <- expand.grid()
```


$$
\begin{aligned}
var(GG_2) &= E[G_2]^2Var(G) + E[G]^2Var(G_2) + Var(G)Var(G_1) \\
          &= (2p_1)^2(2p(1-p)) + (2p)^2(2p_1(1-p_1)) + (2p(1-p))(2p_1(1-p_1)) \\
          &= 8p_1^2p(1-p) + 8p^2p_1(1-p_1) + 4p(1-p)p_1(1-p_1)
\end{aligned}
$$

Check $Var(G_1 G_2)$

```{r}
fn <- function(p1, p2, n=10000)
{
  g1 <- rbinom(10000, 2, p1)
  g2 <- rbinom(10000, 2, p2)
  a=8*p2^2*p1*(1-p1) + 8*p1^2*p2*(1-p2) + 4*p1*(1-p1)*p2*(1-p2)
  b=var(g1*g2)
  return(c(a,b))
}
expand.grid(p1=seq(0.1,0.9,by=0.1), p2=seq(0.1,0.9,by=0.1)) %>%
  {sapply(1:nrow(.), function(i) do.call(fn, .[i,]))} %>%
  t() %>%
  plot
```



## Difference in MAF and LD can have an impact on power

For equal sample size. How is power determined? Relates to R2 and sample size... Expected SE in exposure and outcome

```{r}
res <- res %>% select(-c(rsq, beta))
res <- left_join(res, select(res_eur, rsidcode, pop, rsq, beta), by=c("rsidcode", "pop"))
res <- res %>%
  mutate(
    rsq = TwoSampleMR::get_r_from_pn(pval, n)^2,
    beta = sqrt(rsq / (2*maf*(1-maf))/r2),
    Fstat = rsq * (n-2)/(1-rsq)
  )

head(simulateGP::expected_se(res$beta * res$r2, res$af, res$n, 1))
```


## Difference in MAF and LD can have an impact on weak instrument bias

Use 1/Fstat as the estimate for weak instrument bias. 

```{r}
ratios %>%
  filter(Fstat_ratio > quantile(Fstat_ratio, 0.05) & Fstat_ratio < quantile(Fstat_ratio, 0.95)) %>%
  ggplot(., aes(y=Fstat_ratio, x=pop)) +
    geom_violin() +
    scale_y_log10()
```

```{r}
res2 <- res2 %>% 
  mutate(Fstat = (rsq * n / (1-rsq)))
res2 %>%
  filter(Fstat > quantile(Fstat, 0.05) & Fstat < quantile(Fstat, 0.95)) %>%
ggplot(., aes(y=1/Fstat, x=pop)) +
  scale_y_log10() +
  geom_violin()
```

Non-EUR tend to have greater weak instrument bias. Slightly weird distribution - sometimes lower WIP in non-EUR

Plot against EUR instead

```{r}
temp <- left_join(subset(res2, pop != "EUR"), subset(res2, pop=="EUR", select=c(beta, rsq, Fstat, rsidcode)), by="rsidcode")
temp %>%
  filter(Fstat.x > 1) %>%
  ggplot(., aes(x=1/Fstat.y, y=1/Fstat.x)) +
  geom_point(alpha=0.1) + 
  geom_abline(b=1, colour="white") +
  geom_smooth() +
  facet_wrap(~ pop, scale="free_y") +
  # scale_y_log10() +
  labs(x="Weak instrument bias (EUR)", y="Weak instrument bias (Other pops)")
```

```{r}
temp %>%
  filter(Fstat.x > 1) %>%
  ggplot(., aes(x=1/Fstat.y, y=1/Fstat.x)) +
#  geom_point(alpha=0.1) + 
  geom_abline(b=1, colour="white") +
  geom_smooth(aes(colour=pop)) +
  # scale_y_log10() +
  labs(x="Weak instrument bias (EUR)", y="Weak instrument bias (Other pops)")

```


```{r}
tt <- temp %>% 
  ungroup() %>%
  mutate(fbin = cut(Fstat.y, breaks=30))
tt
ggplot(tt, aes(x=Fstat.y, y=Fstat.x)) +
#  geom_boxplot(outliers = FALSE, aes(group=fbin)) +
  geom_smooth(method="lm", aes(x=Fstat.y)) +
  facet_wrap(~ pop, scales="free") +
  geom_abline(slope=1, intercept=0)
```


Get a slope for these relationships

```{r}
summary(lm(I(1/Fstat.x) ~ I(1/Fstat.y), temp))

group_by(temp, pop) %>%
  do({
    summary(lm(I(1/Fstat.x) ~ I(1/Fstat.y), data=.))$coef %>% as_tibble %>% slice_tail(n=1)
  })
```

So weak instrument bias is about 3x higher in AFR than Europeans assuming same sample size.

```{r}
table(temp$rsq.x > temp$rsq.y)
```

How much of this is allele frequency versus r2?

```{r}
res_eur2 <- subset(res, pop=="EUR") %>%
  mutate(
    rsq = TwoSampleMR::get_r_from_pn(pval, n)^2,
    beta = sqrt(rsq / (2*maf*(1-maf))),
    Fstat = rsq * (n-2)/(1-rsq)
  )

res3 <- left_join(res, dplyr::select(ungroup(res_eur2), rsidcode, beta), by="rsidcode") %>%
  mutate(
    rsq = beta^2 * 2 * maf * (1-maf),
    Fstat = (rsq * n / (1-rsq))
  )
temp2 <- left_join(subset(res3, pop != "EUR"), subset(res3, pop=="EUR", select=c(beta, rsq, Fstat, rsidcode)), by="rsidcode")
temp2 %>%
  filter(Fstat.x > 1) %>%
  ggplot(., aes(x=1/Fstat.y, y=1/Fstat.x)) +
  geom_point() + 
  geom_abline(b=1, colour="white") +
  geom_smooth() +
  facet_wrap(~ pop)
```

```{r}
group_by(temp2, pop) %>%
  do({summary(lm(I(1/Fstat.x) ~ I(1/Fstat.y), data=.))$coef %>% as_tibble %>% slice_tail(n=1)})
```

```{r}
bind_rows(
  temp %>% mutate(what="AF + LD"),
  temp2 %>% mutate(what="AF")
) %>%
  filter(Fstat.x > 1) %>%
  ggplot(., aes(x=1/Fstat.y, y=1/Fstat.x)) +
#  geom_point(alpha=0.1) + 
  geom_abline(b=1, colour="black") +
  geom_smooth(method="lm", aes(colour=pop)) +
  facet_grid(. ~ what) +
  labs(x="Weak instrument bias (EUR)", y="Weak instrument bias (Other pops)") +
  scale_colour_brewer(type="qual")
```


```{r, fig.width=7, fig.height=3}
bind_rows(
  temp %>% mutate(what="AF + LD"),
  temp2 %>% mutate(what="AF")
) %>%
  group_by(pop, what) %>%
    do({summary(lm(I(1/Fstat.x) ~ I(1/Fstat.y), data=.))$coef %>% as_tibble %>% slice_tail(n=1)}) %>%
  {ggplot(., aes(x=pop, y=Estimate)) +
      geom_point(aes(colour=what), position=position_dodge(width=0.5)) +
      geom_errorbar(aes(ymin=Estimate - `Std. Error` * 1.96, ymax = Estimate + `Std. Error` * 1.96, colour=what), position=position_dodge(width=0.5), width=0) +
      labs(colour="Source", y="Weak instrument bias increase\nrelative to EUR discovery", x="Population")
  }
```

The difference in 1/F would be ~2 for Africans without the LD problem (i.e. if you introduce fine mapping effectively).

Quick look at r2 distributions to check they're ok

```{r}
res2 %>%
  ggplot(., aes(x=pop, y=rsq)) +
  geom_violin()
```

```{r}
subset(res2, rsq > 0.01)
```


```{r}
temp <- subset(res2, rsq > 0.01)$rsidcode
rbind(subset(res2, rsq > 0.01), subset(res2, pop=="EUR" & rsidcode %in% temp)) %>% arrange(rsidcode)
```

Rsq in non-EUR higher when the MAF in EUR low.



## Figures

```{r}
 ggplot(NULL, aes(c(-3,3))) +
  geom_area(stat = "function", fun = function(x) {dnorm(x, mean=0.2)}, fill = "#00998a", xlim = c(-3, 1)) +
  geom_area(stat = "function", fun = function(x) {dnorm(x, mean=0.2)}, fill = "grey80", xlim = c(1, 3)) +
  geom_vline(xintercept=0.2) +
  labs(x = "z", y = "", title="Population A") +
  scale_y_continuous(breaks = NULL) +
  scale_x_continuous(breaks = 1)
```

```{r}
 ggplot(NULL, aes(c(-3,3))) +
  geom_area(stat = "function", fun = function(x) {dnorm(x, mean=-0.4)}, fill = "#00998a", xlim = c(-3, 2)) +
  geom_vline(xintercept=-0.4) +
  geom_area(stat = "function", fun = function(x) {dnorm(x, mean=-0.4)}, fill = "grey80", xlim = c(2, 3)) +
  labs(x = "z", y = "", title="Population B") +
  scale_y_continuous(breaks = NULL) +
  scale_x_continuous(breaks = 2)
```

### LD

```{r}
library(ieugwasr)
a <- associations("1:1000000-1150000", "ieu-a-7")
eur_ld <- ieugwasr::ld_matrix(a$rsid, pop="EUR")
afr_ld <- ieugwasr::ld_matrix(a$rsid, pop="AFR")
keepnames <- intersect(colnames(eur_ld), colnames(afr_ld))
rsid_keep <- strsplit(keepnames, split="_") %>% sapply(., function(x) x[1])
d <- subset(a, rsid %in% rsid_keep, select=c(rsid, position))
eur_ld <- eur_ld[colnames(eur_ld) %in% keepnames, colnames(eur_ld) %in% keepnames]
afr_ld <- afr_ld[colnames(afr_ld) %in% keepnames, colnames(afr_ld) %in% keepnames]
# eur_ld[upper.tri(eur_ld)] <- NA
# afr_ld[upper.tri(afr_ld)] <- NA
rownames(eur_ld) <- colnames(eur_ld) <- d$position
rownames(afr_ld) <- colnames(afr_ld) <- d$position
eur_long <- reshape2::melt(eur_ld)
afr_long <- reshape2::melt(afr_ld)
ggplot(eur_long, aes(x=as.factor(Var1), y=as.factor(Var2))) +
  geom_tile(aes(fill=value^2))

eur_long$value[eur_long$Var1 > eur_long$Var2] <- 0
afr_long$value[afr_long$Var1 > afr_long$Var2] <- 0
geur <- ggplot(eur_long, aes(x=as.factor(Var1), y=as.factor(Var2))) +
  geom_tile(aes(fill=abs(value))) +
  scale_fill_gradient(low="white", high="red") +
  theme(axis.text=element_blank(), axis.ticks=element_blank(), legend.position = "none") +
  labs(x="", y="")
ggsave(geur, file="geur.png", width=6, height=6)

gafr <- ggplot(afr_long, aes(x=as.factor(Var1), y=as.factor(Var2))) +
  geom_tile(aes(fill=abs(value))) +
  scale_fill_gradient(low="white", high="red") +
  theme(axis.text=element_blank(), axis.ticks=element_blank(), legend.position = "none") +
  labs(x="", y="")
ggsave(gafr, file="gafr.png", width=6, height=6)
```


```{r}
library("grid")

gafr <- gafr + annotation_custom(
  grob = textGrob(label = "AFR", rot = -45,
    x = unit(1.1, "npc"), y = unit(1.1, "npc")))

print(gafr, vp = viewport(width = unit(0.5, "npc"),
  height = unit(0.5, "npc"), angle = 45))
```

















