---
title: "Probability of a random variable being larger than all other random variables in a multivariate normal vector"
output: html_notebook
---

I have 1k SNPs in a region. I know the causal variant and the LD matrix. The effect size at each SNP will be related to the allele frequency and the LD at all other variants. The SE across the SNPs will be correlated in relation to the LD. I can generate the expected effect size and the variance covariance matrix of the effects. Once I have that, I can generate beta values from a multivariate normal distribution, and determine how often each of the SNPs is the top SNP.

Is there a faster way to do this by getting the probability from a multivariate normal distribution?

Related to this question: https://stats.stackexchange.com/a/4181


```{r}
install.packages("MCMCpack")
library(MCMCpack)
library(dplyr)
library(simulateGP)
library(MASS)
library(mvtnorm)
```


Empirical simulation for probabilities, case of 3 variables

```{r}
n <- 3
mu <- rnorm(n)
S <- rwish(n, diag(n))
emp <- mvrnorm(10000, mu, S)
res <- apply(emp, 1, function(x) which.max(x)) %>% table() %>% prop.table()
res
```


```{r}
A <- matrix(c(1,-1,0, 1,0,-1), nrow = 2, byrow = TRUE)
newMu <- as.vector(A %*% mu)
newS <- A %*% S %*% t(A)
pmvnorm(lower=c(0,0), mean = newMu, sigma = newS)

A <- matrix(c(1,-1,0, 1,0,-1), nrow = 2, byrow = TRUE)
A <- A[,c(2,1,3)]
newMu <- as.vector(A %*% mu)
newS <- A %*% S %*% t(A)
pmvnorm(lower=c(0,0), mean = newMu, sigma = newS)

A <- matrix(c(1,-1,0, 1,0,-1), nrow = 2, byrow = TRUE)
A <- A[,c(2,3,1)]
newMu <- as.vector(A %*% mu)
newS <- A %*% S %*% t(A)
pmvnorm(lower=c(0,0), mean = newMu, sigma = newS)
```

Increase to arbitrary variables

```{r}
swap_1 <- function(n, i)
{
  ind <- 1:n
  if(i == 1) return(ind)
  ind[i] <- 1
  ind[1:(i-1)] <- 2:i
  return(ind)
}
#sapply(1:7, function(i) swap_1(7, i))

n <- 1000
mu <- rnorm(n)
S <- rwish(n, diag(n))
emp <- mvrnorm(10000, mu, S)
res <- apply(emp, 1, function(x) which.max(x)) %>% table() %>% prop.table()
res

A <- cbind(
  rep(1, n-1),
  diag(rep(-1, n-1))
) %>% as.matrix()

emp <- sapply(1:n, function(i)
{
  A <- A[,swap_1(n,i)]
  newMu <- as.vector(A %*% mu)
  newS <- A %*% S %*% t(A)
  pmvnorm(lower=rep(0,n-1), mean = newMu, sigma = newS)
})
plot(emp ~ as.numeric(res))
```


Theoretical result works fine but is slower than empirical sampling.

