---
title: Analysis of EUR vs EAS traits
author: Gibran Hemani
---

```{r}
library(knitr)
knitr::opts_chunk$set(cache=TRUE)
```

```{r}
library(CAMeRa)
library(ggplot2)
library(ggstance)
library(dplyr)
library(here)
library(ieugwasr)
```


```{r}
fn <- readRDS(here("data", "bbj", "fn.rds"))
res <- readRDS(here("data", "bbj", "res.rds"))
length(res)
names(res)
head(fn)
sets <- 1:length(res)
```

## MR results

```{r}
mrres <- lapply(sets, \(i) {
    res[[i]]$mrres %>% 
    mutate(
        exposure=fn$exposure[i], 
        outcome=fn$outcome[i],
        eur_id_x = fn$eur_id_x[i],
        eas_id_x = fn$eas_id_x[i],
        eur_id_y = fn$eur_id_y[i],
        eas_id_y = fn$eas_id_y[i],
    )
}) %>% bind_rows()
```

Any heterogeneity?

```{r}
mrres %>% 
    filter(instrument=="FEMA" & eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52") %>%
    filter(Qjpval < 0.05)
```

```{r}
mrres %>% 
    filter(instrument=="FEMA" & eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52") %>%
    filter(pops == "All") %>%
    mutate(Qfdr = p.adjust(Qjpval, "fdr")) %>% as.data.frame
    filter(Qfdr < 0.05)
```


Not really after multiple testing correction.


```{r}
mrres %>% 
    filter(instrument=="Raw" & eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52") %>%
    filter(pops == "All") %>%
    mutate(Qfdr = p.adjust(Qjpval, "fdr")) %>%
    filter(Qfdr < 0.05)
```




```{r}
mrres %>% 
    filter(instrument=="Raw" & eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52") %>%
    filter(pops == "All") %>% mutate(Qfdr = p.adjust(Qjpval)) %>% filter(Qfdr < 0.05) %>% str
```


```{r}
mrres %>%
    filter(instrument=="FEMA" & eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52") %>% 
    mutate(lci = Estimate - 1.96 * `Std. Error`, uci = Estimate + 1.96 * `Std. Error`) %>%
    mutate(lci = pmax(-0.6, lci), uci = pmin(1.1, uci)) %>%
    {
    ggplot(., aes(x=Estimate, y=pops)) +
    geom_rect(data=subset(., pops == "All"), aes(alpha = Qjpval < 0.05), xmin = -Inf,xmax = Inf, ymin = -Inf, ymax = Inf, fill="yellow") +
    geom_vline(xintercept=0, linetype="dotted") +
    geom_point(aes(colour=pops=="All")) +
    geom_errorbarh(aes(colour=pops=="All", xmin=lci, xmax=uci), height=0) + 
    facet_grid(outcome ~ exposure) +
    theme(strip.text.y=element_text(angle=0), legend.position="none") +
    scale_alpha_manual(values=c(0, 0.2)) +
    xlim(-0.6, 1.1) +
    labs(y="")
}
ggsave(here("results", "bbj_mrres.pdf"), width=10, height=8)
```


```{r}
mrres %>%
    filter(instrument=="Raw" & eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52") %>% 
    mutate(lci = Estimate - 1.96 * `Std. Error`, uci = Estimate + 1.96 * `Std. Error`) %>%
    mutate(lci = pmax(-0.6, lci), uci = pmin(1.1, uci)) %>%
    {
    ggplot(., aes(x=Estimate, y=pops)) +
    geom_rect(data=subset(., pops == "All"), aes(alpha = Qjpval < 0.05), xmin = -Inf,xmax = Inf, ymin = -Inf, ymax = Inf, fill="yellow") +
    geom_vline(xintercept=0, linetype="dotted") +
    geom_point(aes(colour=pops=="All")) +
    geom_errorbarh(aes(colour=pops=="All", xmin=lci, xmax=uci), height=0) + 
    facet_grid(outcome ~ exposure) +
    theme(strip.text.y=element_text(angle=0), legend.position="none") +
    scale_alpha_manual(values=c(0, 0.2)) +
    xlim(-0.6, 1.1) +
    labs(y="")
}
ggsave(here("results", "bbj_mrres_raw.pdf"), width=10, height=8)
```














## Simulation to check interpretation

T2D in East Asians onset occurs at a lower BMI threshold. What should this do to the effect of BMI on T2D?

```{r}
n <- 100000
bmi <- rnorm(n, 22, 5)
l1 <- rep(0.02, n)
l1[bmi > 26] <- bmi[bmi > 26] * 0.01
l2 <- rep(0.02, n)
l2[bmi > 23] <- bmi[bmi > 23] * 0.01
t2d1 <- rbinom(n, 1, plogis(l1))
t2d2 <- rbinom(n, 1, plogis(l2))
table(t2d1)
table(t2d2)

summary(glm(t2d1 ~ bmi, family="binomial"))
summary(glm(t2d2 ~ bmi, family="binomial"))
```

- Higher starting point leads to higher effect size
- That means that the slope must be smaller. Effect size is much more sensitive to slope than to starting point 



```{r}
adip1 <- rnorm(n, 0, 5)
adip2 <- rnorm(n, 0, 5)
bmi1 <- adip1 * 2 + rnorm(n, 22, 5)
bmi2 <- adip2 * 5 + rnorm(n, 22, 5)
t2d1 <- rbinom(n, 1, plogis(0.02 * adip1))
t2d2 <- rbinom(n, 1, plogis(0.02 * adip2))

summary(glm(t2d1 ~ bmi1, family="binomial"))
summary(glm(t2d2 ~ bmi2, family="binomial"))


tapply(bmi1, t2d1, mean)
tapply(bmi2, t2d2, mean)

```




## Reanalyse BMI and T2D

```{r}
load(here("data/bbj/examples_bmi_t2d.rdata"))

x$instrument_raw %>% head

library(TwoSampleMR)

dat <- make_dat("ieu-b-40", c("ebi-a-GCST006867", "ebi-a-GCST010118"))
a <- mr(dat)
```


```{r}
subset(mrres, exposure == "Body Mass Index" & outcome == "Type 2 Diabetes")
```


```{r}
md <- make_dat("ieu-b-40", "ukb-e-250_CSA")
mr(md)
```


```{r}
adi <- make_dat("ukb-b-16407", c("ieu-b-40", "bbj-a-1"))
mr(adi, method_list="mr_ivw")
```


```{r}
adi2 <- make_dat("ukb-b-16407", c("ebi-a-GCST006867", "ebi-a-GCST010118", "ukb-e-250_CSA"))
mr(adi2, method_list="mr_ivw")
```



Naive estimates
Eur exposure -> Eur outcome
Eur exposure -> EAS outcome

```{r}
analysis_plan <- filter(mrres, pops == "All") %>%
    filter(instrument=="FEMA" & eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52") %>%
    group_by(exposure, outcome) %>% do({
    tibble(
        id_x = .$eur_id_x[1],
        id_y = c(.$eur_id_y[1], .$eas_id_y[1]),
        pop = c("EUR", "EAS")
    )
})

naive_dat <- group_by(analysis_plan, exposure, outcome) %>% do({
    X <- .
    dat <- make_dat(unique(X$id_x), unique(X$id_y))
    dat
})


naive_dat2 <- subset(naive_dat, paste(id.exposure, id.outcome) %in% paste(analysis_plan$id_x, analysis_plan$id_y))
dim(naive_dat2)
dim(naive_dat)



naive_res <- group_by(analysis_plan, exposure, outcome) %>% do({
    X <- .
    dat <- subset(naive_dat, id.exposure == X$id_x & id.outcome %in% X$id_y)
    a <- mr(dat, method_list=c("mr_wald_ratio", "mr_ivw"))
    fe <- fixed_effects_meta_analysis(a$b, a$se)
    a %>% select(id.exposure, id.outcome, nsnp, b, se, pval)
    a$Qj <- fe$Qj
    a$Qjpval <- fe$Qjpval
    a$pops <- c("EUR", "EAS")
    fe <- tibble(b=fe$beta, se=fe$se, pval=fe$pval, Qj=fe$Q, Qjpval=fe$Qpval, pops="All")
    bind_rows(fe, a) %>% select(-c(exposure, outcome)) %>% mutate(instrument="Naive")
})

naive_res %>% filter(pops == "All") %>%
    mutate(Qfdr = p.adjust(Qjpval, "fdr")) %>%
    filter(Qfdr < 0.05)

    
    a <- subset(X, select=c(id_y, pop)) %>% left_join(., a, by=c("id_y"="id.outcome"))
    fe <- fixed_effects_meta_analysis(a$b, a$se)
    a$Q
    return(a)

```


```{r}
o <- mrres %>% filter(eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52") %>% 
    select(pops, b=Estimate, se=`Std. Error`, pval=`Pr(>|t|)`, Qj, Qjpval, exposure, outcome, instrument) %>% bind_rows(naive_res)
o %>% filter(pops=="All") %>% group_by(instrument) %>% mutate(Qfdr = p.adjust(Qjpval, "fdr")) %>% summarise(nsig=sum(Qfdr < 0.05))
```


```{r}
o %>% filter(instrument == "Naive") %>%
ggplot(aes(x=b, y=pops)) +
    facet_grid(outcome ~ exposure) +
    geom_point(aes(colour=instrument)) +
    geom_vline(xintercept=0, linetype="dotted") +
    geom_errorbarh(aes(xmin=b - se * 1.96, xmax=b + se * 1.96, colour=instrument), height=0) +
    geom_vline(xintercept=0, linetype="dotted") +
    facet_grid(outcome ~ exposure) +
    theme(strip.text.y=element_text(angle=0), legend.position="none") +
    labs(y="")
```


```{r}
o %>%
    filter(instrument=="Naive") %>% 
    mutate(lci = b - 1.96 * se, uci = b + 1.96 * se) %>%
    mutate(lci = pmax(-0.6, lci), uci = pmin(1.1, uci)) %>%
    {
    ggplot(., aes(x=b, y=pops)) +
    geom_rect(data=subset(., pops == "All"), aes(alpha = Qjpval < 0.05), xmin = -Inf,xmax = Inf, ymin = -Inf, ymax = Inf, fill="yellow") +
    geom_vline(xintercept=0, linetype="dotted") +
    geom_point(aes(colour=instrument), position=ggstance::position_dodgev(0.3)) +
    geom_errorbarh(aes(colour=instrument, xmin=lci, xmax=uci), height=0, position=ggstance::position_dodgev(0.3)) + 
    facet_grid(outcome ~ exposure) +
    theme(strip.text.y=element_text(angle=0), legend.position="bottom") +
    scale_alpha_manual(values=c(0, 0.2)) +
    xlim(-0.6, 1.1) +
    labs(y="") +
    scale_colour_brewer(type="qual")
}
ggsave(here("results", "bbj_mrres_naive.pdf"), width=10, height=8)
```


```{r}
o %>%
    # filter(instrument=="Naive") %>% 
    mutate(lci = b - 1.96 * se, uci = b + 1.96 * se) %>%
    mutate(lci = pmax(-0.6, lci), uci = pmin(1.1, uci)) %>%
    {
    ggplot(., aes(x=b, y=pops)) +
    geom_rect(data=subset(., pops == "All"), aes(alpha = Qjpval < 0.05), xmin = -Inf,xmax = Inf, ymin = -Inf, ymax = Inf, fill="yellow") +
    geom_vline(xintercept=0, linetype="dotted") +
    geom_point(aes(colour=instrument), position=ggstance::position_dodgev(0.3)) +
    geom_errorbarh(aes(colour=instrument, xmin=lci, xmax=uci), height=0, position=ggstance::position_dodgev(0.3)) + 
    facet_grid(outcome ~ exposure) +
    theme(strip.text.y=element_text(angle=0), legend.position="bottom") +
    scale_alpha_manual(values=c(0, 0.2)) +
    xlim(-0.6, 1.1) +
    labs(y="") +
    scale_colour_brewer(type="qual")
}
ggsave(here("results", "bbj_mrres_all.pdf"), width=10, height=8)
```




```{r}
subset(mrres, pops == "All" & instrument == "FEMA") %>% mutate(Qfdr = p.adjust(Qjpval, "fdr")) %>% filter(Qfdr < 0.05) %>% {.$f}
```


Power increase

```{r}
mrres %>%
    filter(instrument=="FEMA" & eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52") %>% group_by(exposure, outcome) %>% arrange(`Std. Error`) %>% slice(1) %>% ungroup() %>% {table(.$pops)}
```

```{r}
mrres %>%
    filter(instrument=="FEMA" & eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52") %>% group_by(exposure, outcome) %>% arrange(`Std. Error`) %>% filter(pops != "EAS") %>% slice(1) %>% ungroup() %>% {table(.$pops)}
```

```{r}
mrres %>%
    filter(instrument=="FEMA" & eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52") %>% 
    group_by(exposure, outcome) %>% filter(pops != "EUR") %>% filter(`Pr(>|t|)` < 0.05 & Qjpval > 0.05) %>% select(pops, exposure, outcome)  %>% as.data.frame
```

## Instrument agreement

```{r}
fn$index <- 1:nrow(fn)
fn2 <- fn %>% filter(eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52")
fn2 <- fn2 %>% filter(!duplicated(exposure))
fn2
sets <- fn2$index
temp <- lapply(sets, \(i) {
    res[[i]]$instrument_specificity %>% mutate(
        exposure = fn$exposure[i]
    )
}) %>% bind_rows() %>% filter(metric == "Sign")

inner_join(subset(temp, datum=="Expected"), subset(temp, datum == "Observed"), by=c("discovery", "replication", "exposure", "instrument")) %>%
    ggplot(., aes(x=value.x, y=value.y)) +
    geom_point(aes(colour=paste(exposure), shape=instrument)) +
    geom_smooth(method="lm", aes(colour=exposure, group=discovery), se=FALSE) +
    # facet_wrap(~ paste(discovery), scale="free") +
    geom_abline(slope=1, intercept=0, linetype="dotted") +
    labs(x="Expected instrument sign agreement", y="Observed instrument sign agreement", colour="Ancestry", shape="Outcome") +
    scale_colour_brewer(type="qual", palette=3)
```

## Instrument heterogeneity

```{r}
fn$index <- 1:nrow(fn)
fn2 <- fn %>% filter(eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52")
fn2 <- fn2 %>% filter(!duplicated(exposure))
fn2
sets <- fn2$index
temp <- lapply(sets, \(i) {
    res[[i]]$instrument_heterogeneity %>% mutate(
        exposure = fn$exposure[i]
    )
}) %>% bind_rows() %>% filter(exposure != "Smoking Initiation")
temp
temp$disc <- "EUR"
temp$disc[temp$Reference %in% fn$eas_id_x] <- "EAS"
temp$rep <- "EUR"
temp$rep[temp$Replication %in% fn$eas_id_x] <- "EAS"

temp %>% ggplot(., aes(x=exposure, y=agreement)) +
geom_point(aes(colour=instrument), position=position_dodge(width=0.1)) +
geom_errorbar(aes(colour=instrument, ymin=agreement-se*1.96, ymax=agreement+se*1.96), width=0, position=position_dodge(width=0.1)) +
facet_grid(disc ~ .)
```

```{r}
temp %>% as.data.frame
```


```{r}
xy <- lapply(sets, \(i) {
    load(fn$newpath[i])
    x$harmonise(exp=x$instrument_raw)
    a1 <- x$harmonised_dat %>% mutate(instrument="Raw")
    x$harmonise(exp=x$instrument_fema)
    a2 <- x$harmonised_dat %>% mutate(instrument="FEMA")
    a <- bind_rows(
        inner_join(
            a1 %>% filter(pops == "EUR") %>% select(SNP, EUR=beta.x, EURse=se.x, instrument),
            a1 %>% filter(pops == "EAS") %>% select(SNP, EAS=beta.x, EASse=se.x, instrument),
        ),
        inner_join(
            a2 %>% filter(pops == "EUR") %>% select(SNP, EUR=beta.x, EURse=se.x, instrument),
            a2 %>% filter(pops == "EAS") %>% select(SNP, EAS=beta.x, EASse=se.x, instrument),
        )
    ) %>% mutate(exposure = fn$exposure[i])
    a
}) %>% bind_rows()

ggplot(xy %>% filter(!grepl("Smoking Initiation", exposure)), aes(x=EUR, y=EAS)) +
geom_point(aes(colour=instrument), alpha=0.2) +
geom_errorbarh(aes(xmin=EUR - EURse * 1.96, xmax=EUR + EURse * 1.96, colour=instrument), height=0, alpha=0.2) +
geom_errorbar(aes(ymin=EAS - EASse * 1.96, ymax=EAS + EASse * 1.96, colour=instrument), width=0, alpha=0.2) +
geom_abline(slope=1, intercept=0) +
geom_smooth(method="lm", aes(colour=instrument)) +
facet_wrap(~ exposure)
```


## Pleiotropy outliers

```{r}
fn2 <- fn %>% filter(eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52")
fn2
sets <- fn2$index

temp <- lapply(sets, \(i) {
    if(is.null(res[[i]]$pleiotropy)) return(NULL)
    res[[i]]$pleiotropy %>% mutate(
        exposure = fn$exposure[i],
        outcome = fn$outcome[i]
    )
}) %>% bind_rows() %>% filter(metric == "Sign")
temp

inner_join(subset(temp, datum=="Expected"), subset(temp, datum == "Observed"), by=c("disc", "rep", "exposure", "outcome")) %>%
    {summary(lm(value.y ~ value.x, data=.))}

inner_join(subset(temp, datum=="Expected"), subset(temp, datum == "Observed"), by=c("disc", "rep", "exposure", "outcome")) %>%
    ggplot(., aes(x=value.x, y=value.y)) +
    geom_point(aes(colour=disc)) +
    geom_smooth(method="lm", aes(colour=disc, group=disc), se=FALSE) +
    # facet_wrap(~ paste(discovery), scale="free") +
    geom_abline(slope=1, intercept=0, linetype="dotted") +
    labs(x="Expected outlier direction agreement", y="Observed outlier direction agreement", colour="Discovery ancestry", shape="Outcome") +
    scale_colour_brewer(type="qual", palette=3)
```


## MR GxE

```{r}
fn2 <- fn %>% filter(eur_id_x != "ieu-b-38" & eur_id_y != "ieu-b-38" & outcome != "Schizophrenia" & eas_id_y != "bbj-a-52")
fn2
sets <- fn2$index
temp <- lapply(sets, \(i) {
    if(is.null(res[[i]]$mrgxe)) return(NULL)
    res[[i]]$mrgxe %>% mutate(
        exposure = fn$exposure[i],
        outcome = fn$outcome[i],
        index = i
    )
}) %>% bind_rows()
temp
```


```{r}
filter(temp, p.adjust(pval, "fdr") < 0.05)
filter(temp, p.adjust(Qpval, "fdr") < 0.05)
```


```{r}
o <- filter(temp, p.adjust(Qpval, "fdr") < 0.05)

tt <- lapply(o$index, \(i) {
    load(fn$newpath[i])
    x$harmonise(exp=x$instrument_fema)
    x$estimate_instrument_heterogeneity_per_variant()
    x$mrgxe()
    osnp <- x$instrument_heterogeneity_per_variant %>% filter(Qfdr < 0.05) %>% {.$SNP}
    qsnp <- x$mrgxe_res %>% mutate(fdr=p.adjust(a_pval, "fdr")) %>% filter(fdr < 0.05) %>% {.$SNP}
    subset(x$harmonised_dat, SNP %in% osnp) %>% mutate(sig = SNP %in% qsnp, exposure = fn$exposure[i], outcome=fn$outcome[i])
}) %>% bind_rows()

ggplot(tt, aes(x=beta.x, y=beta.y)) +
    geom_errorbarh(aes(xmin=beta.x - se.x * 1.96, xmax=beta.x + se.x * 1.96), height=0, alpha=0.2) +
    geom_errorbar(aes(ymin=beta.y - se.y * 1.96, ymax=beta.y + se.y * 1.96), width=0, alpha=0.2) +
    geom_abline(slope=1, intercept=0) +
    geom_point(aes(colour=pops)) +
    geom_smooth(se=FALSE, method="lm", aes(group=SNP, colour=sig)) +
    geom_smooth(se=FALSE, method="lm", aes(colour=pops)) +
    labs(colour="MR GxE FDR < 0.05") +
    facet_wrap(~ paste(outcome, "~", exposure)) +
    scale_colour_brewer(type="qual")
```

```{r}
subset(tt, exposure == "Body Mass Index" & sig)$SNP %>% unique
```

10:114758349 - TCF7L2
12:14413931 - ATF7IP
15:51748610 - DMXL2
18:58039276 - MC4R
3:85819412 - CADM2
3:185499057 - IG2BP2
6:20694884 - CDKAL1

