---
title: UKB T2D by adiposty and ancestry
---


```{r}
library(dplyr)
library(data.table)
library(ggplot2)
library(tidyr)
library(TwoSampleMR)
```



## Organise data

codes:

- trunk fat percentage = 23127
- body fat percentage = 23099
- body mass index = 21001

```{r}
case <- fread("t2dcases.csv") %>% mutate(t2d=1)
cont <- fread("t2dcontrols.csv") %>% mutate(t2d=0)

anc <- fread("ukb_InferredAncestry.txt") %>% as_tibble() %>% select(eid=FID, anc=Ancestry)

dat <- rbind(case, cont) %>% left_join(anc, by="eid")
dat <- subset(dat, !grepl(";", anc)) %>% filter(!is.na(anc))

names(dat) <- c("eid", "tfp", "bfp", "bmi", "sex", "age", "age2", "t2d", "anc")
```



```{r}
table(dat$anc)
```


The effect of TFP on BMI is different by ancestry:

```{r}
dat %>% group_by(anc) %>% summarise(b_tfp_bmi = lm(bmi ~ tfp)$coef[2])
dat
```

The response of BMI to TFP is about 25% lower in EAS than EUR. Get the MR estimate:

```{r}
adi <- make_dat("ukb-b-16407", c("ieu-b-40", "bbj-a-1"))
mr(adi, method_list="mr_ivw")
```

The response of BMI to TFP is substantially lower in BBJ than in EUR.

What is the influence of TFP on T2D? This is a naive estimate, using the same EUR bgx for the EUR and EAS outcome:

```{r}
adi2 <- make_dat("ukb-b-16407", c("ebi-a-GCST006867", "ebi-a-GCST010118", "ukb-e-250_CSA", "ukb-e-250_AFR"))
mr(adi2, method_list="mr_ivw")
```


```{r}
adi3 <- make_dat("ukb-b-8909", c("ebi-a-GCST006867", "ebi-a-GCST010118", "ukb-e-250_CSA", "ukb-e-250_AFR"))
mr(adi3, method_list="mr_ivw")

```


```{r}
dat %>% filter(
    bmi < 40
) %>% 
    ggplot(aes(x=tfp, y=bmi, color=anc)) + geom_smooth(se=FALSE) +
    scale_colour_brewer(type="qual")
```

EAS and SAS have higher TFP for a given BMI

```{r}
dat %>% mutate(tfp_bins = round(tfp)) %>%
    group_by(tfp_bins, anc) %>%
    summarise(t2d_prev = mean(t2d, na.rm=T)) %>%
    ggplot(aes(x=tfp_bins, y=t2d_prev, color=anc)) + 
        geom_smooth(se=FALSE) +
        scale_colour_brewer(type="qual")
```


```{r}

ggplot(dat, aes(x=tfp, y=t2d, color=anc)) + 
    geom_smooth() +
    scale_colour_brewer(type="qual")

```



```{r}
ggplot(dat, aes(x=bmi, y=t2d, color=anc)) + 
    geom_smooth() +
    scale_colour_brewer(type="qual")
```


```{r}
group_by(dat, anc) %>%
    summarise(t2d_prev = mean(t2d, na.rm=T))
```



```{r}
ggplot(dat, aes(x=bmi)) +
geom_density(aes(fill=anc), alpha=0.5) +
scale_fill_brewer(type="qual")
```


```{r}
ggplot(dat, aes(x=tfp)) +
geom_density(aes(fill=anc), alpha=0.5) +
scale_fill_brewer(type="qual")
```

```{r}
ggplot(dat, aes(x=bfp)) +
geom_density(aes(fill=anc), alpha=0.5) +
scale_fill_brewer(type="qual")
```


```{r}
library(tidyr)

dat %>% 
    select(eid, tfp, bfp, bmi, anc) %>%
    tidyr::pivot_longer(-c(eid, anc), names_to="measure") %>%
    group_by(anc, measure) %>%
    summarise(mean = mean(value, na.rm=T), sd = sd(value, na.rm=T)) %>%
    ggplot(aes(y=anc, x=mean, colour=anc)) +
    geom_point() +
    geom_errorbarh(aes(xmin=mean-sd, xmax=mean+sd), height=0.1) +
    facet_grid(measure ~ ., scales="free_x") +
    scale_colour_brewer(type="qual")

group_by(dat, anc) %>%

    summarise(t2d_prev = mean(t2d, na.rm=T))

