---
title: "Pleiotropy simulations"
output: html_notebook
---


```{r}

### Introducing pleiotropy




```{r}
effects1 <- effects2 <- choose_effects(200, 0.4)
effects1[1:50] <- 0
effects2[51:100] <- 0

# create genotypes
g1 <- make_geno(10000, 200, 0.5)
g2 <- make_geno(10000, 200, 0.5)

# create confounder
u1 <- rnorm(10000)
u2 <- rnorm(10000)

# create x
x1 <- make_phen(c(0.1, effects1), cbind(u1, g1))
x2 <- make_phen(c(0.1, effects2), cbind(u2, g2))

plei <- choose_effects(200, 0.4)

# create y
y1 <- make_phen(c(0.1, 0.2, plei), cbind(u1, x1, g1))
y2 <- make_phen(c(0.1, 0.2, plei), cbind(u2, x2, g2))

# Summary data for pop1 
ss1 <- get_effs(x1, y1, g1)

# Summary data for pop2
ss2 <- get_effs(x2, y2, g2)

mres1 <- mr(ss1, method="mr_ivw")
mres2 <- mr(ss2, method="mr_ivw")


plot(beta.outcome ~ beta.exposure, ss1)
abline(a=0, b=1)
abline(a=0, b=mres1$b, col="red")

# Organise the data
dat <- tibble(
	y1 = ss1$beta.outcome,
	y2 = ss2$beta.outcome,
	x1 = ss1$beta.exposure,
	x2 = ss2$beta.exposure
)

dat$xb <- rowMeans(cbind(dat$x1, dat$x2))
dat$xb[1:100] <- 0
dat$x1[1:50] <- 0
dat$x1[101:200] <- 0
dat$x2[51:100] <- 0
dat$x2[101:200] <- 0

model <- '
y1 ~ b_1*x1 + b_1*xb + p_1*x2
y2 ~ p_2*x1 + b_2*xb + b_2*x2
x1 ~~ 0 * x2
x1 ~~ 0 * xb
x2 ~~ 0 * xb
'
cor(dat)
```
```{r}
mod <- sem(model, data=dat)
summary(mod)
```


```{r}
semPaths(mod, edge.label.cex = 0.7, edge.label.position=0.7)
```

```{r}
nsnp <- 200
map <- tibble(
	snp=1:200,
	af=0.5
) %>%
	mutate(beta = 
		case_when(
			snp > 
	))


xp <- generate_gwas_params(map, 0.4)

x1p <- xp %>%
	mutate(beta = 
		case_when(
			snp <= 50 ~ 0,
			TRUE ~ beta
		)
	)

y1p <- x1p %>%
	mutate(beta = beta * 0.2 + 0.1)

x2p <- xp %>%
	mutate(beta = 
		case_when(
			snp > 50 & snp <= 100 ~ 0,
			TRUE ~ beta
		)
	)

y2p <- x2p %>%
	mutate(beta = beta * 0.2 + 0.1)

yp <- xp %>%
	mutate(beta = beta * 0.2)

summary(lm(yp$beta ~ -1 + xp$beta))
summary(lm(y1p$beta ~ -1 + x1p$beta))
summary(lm(y2p$beta ~ -1 + x2p$beta))

y1 <- generate_gwas_ss(y1p, 1000000)
y2 <- generate_gwas_ss(y2p, 1000000)
x1 <- generate_gwas_ss(x1p, 1000000)
x2 <- generate_gwas_ss(x2p, 1000000)

dat <- tibble(
	y1 = y1$bhat,
	y2 = y2$bhat,
	x1 = x1$bhat,
	x2 = x2$bhat
)

dat$xb <- rowMeans(cbind(dat$x1, dat$x2))
dat$xb[1:100] <- 0
dat$x1[1:50] <- 0
dat$x1[101:200] <- 0
dat$x2[51:100] <- 0
dat$x2[101:200] <- 0

summary(lm(cbind(y1, y2) ~ -1 + x1 + x2 + xb, dat))


summary(lm(y1$bhat ~ -1 + x1$bhat))


model <- '
# linear model
y1 ~ b11*x1 + b11*xb + p1*x2
y2 ~ p2*x1 + b22*xb + b22*x2
'

mod <- sem(model, data=dat, meanstructure=FALSE)
summary(mod)


model <- '
# linear model
y1 ~ b11*x1 + b11*xb + p1*x2
y2 ~ p2*x1 + b22*xb + b22*x2
'

mod <- sem(model, data=dat, meanstructure=FALSE)
summary(mod)


dat2 <- tibble(
	w1 = abs(x1$bhat) / y1$se,
	w2 = abs(x2$bhat) / y2$se,
	Y1 = y1$bhat / x1$bhat * w1,
	Y2 = y2$bhat / x2$bhat * w2,
)

summary(lm(Y1 ~ -1 + w1, dat2))
summary(lm(y1$bhat ~ -1 + x1$bhat, weight=1/y1$se^2))





model <- '
y1 ~ xb + x1
'
summary(sem(model, data=dat))



up <- map %>%
	generate_gwas_params(., 0.1)

xp <- up %>%
	mutate(beta=beta * 0.2)

yp <- up %>%
	mutate(beta=beta * 0.4 + beta * 0.3)

x1p <- xp %>%
	mutate(beta = 
		case_when(
			snp <= 50 ~ 0,
			TRUE ~ beta
		)
	)

y1p <- x1p %>%
	mutate(beta = case_when(
		beta == 0 ~ beta * 0.4,
		TRUE ~ beta * 0.4 + beta * 0.3
	))

x2p <- xp %>%
	mutate(beta = 
		case_when(
			snp > 50 & snp <= 100 ~ 0,
			TRUE ~ beta
		)
	)

y2p <- x2p %>%
	mutate(beta = case_when(
		beta == 0 ~ beta * 0.4,
		TRUE ~ beta * 0.4 + beta * 0.3
	))

plot(y1p$beta ~ x1p$beta)

summary(lm(yp$beta ~ xp$beta))

summary(lm(y2p$beta ~ x1p$beta, subset=x1p$beta == 0))



####


# create genotypes
g1 <- make_geno(10000, 200, 0.5)
g2 <- make_geno(10000, 200, 0.5)

# create confounder
plei1 <- plei2 <- choose_effects(200, 0.4)
effects1[1:50] <- 0
effects2[51:100] <- 0
u1 <- make_phen(plei, g1)
u2 <- make_phen(plei, g2)

x1 <- make_phen(0.6, u1)
x2 <- make_phen(0.6, u2)

y1 <- make_phen(c(0.6, -0.4), cbind(u1, x1))
y2 <- make_phen(c(0.6, -0.4), cbind(u2, x2))


# Summary data for pop1 
ss1 <- get_effs(x1, y1, g1)

# Summary data for pop2
ss2 <- get_effs(x2, y2, g2)

(mres1 <- mr(ss1, method="mr_ivw"))
(mres2 <- mr(ss2, method="mr_ivw"))


plot(beta.outcome ~ beta.exposure, ss1)
abline(a=0, b=1)
abline(a=0, b=mres1$b, col="red")

# Organise the data
dat <- tibble(
	y1 = ss1$beta.outcome,
	y2 = ss2$beta.outcome,
	x1 = ss1$beta.exposure,
	x2 = ss2$beta.exposure
)

dat$xb <- rowMeans(cbind(dat$x1, dat$x2))
dat$xb[1:100] <- 0
dat$x1[1:50] <- 0
dat$x1[101:200] <- 0
dat$x2[51:100] <- 0
dat$x2[101:200] <- 0


model <- '
y1 ~ b_1*x1 + b_1*xb + p_1*x2
y2 ~ p_2*x1 + b_2*xb + b_2*x2
'

mod <- sem(model, data=dat)
summary(mod)


########

# try again


# create genotypes
g1 <- make_geno(10000, 50, 0.5)
g2 <- make_geno(10000, 50, 0.5)
gb <- make_geno(10000, 100, 0.5)

eff1 <- choose_effects(50, 0.2)
eff2 <- choose_effects(50, 0.2)
effb <- choose_effects(100, 0.2)
x1 <- make_phen(eff1, g1)
x2 <- make_phen(eff2, g2)
xb <- make_phen(effb, gb)

plei1 <- choose_effects(50, 0.1)
plei2 <- choose_effects(50, 0.1)
y1 <- make_phen(c(eff1, eff2, effb, 0.2, 0.2), cbind(g1, g2, gb, x1, xb))
y2 <- make_phen(c(eff1, eff2, effb, 0.2, 0.2), cbind(g1, g2, gb, x2, xb))




# Summary data for pop1 
ss1 <- get_effs(x1, y1, g1)

# Summary data for pop2
ss2 <- get_effs(x2, y2, g2) %>% mutate(SNP = SNP + 50)

ssb1 <- get_effs(xb, y1, gb) %>% mutate(SNP = SNP + 100)
ssb2 <- get_effs(xb, y2, gb) %>% mutate(SNP = SNP + 100)


mres1 <- mr(ss1, method="mr_ivw")
mres2 <- mr(ss2, method="mr_ivw")


plot(beta.outcome ~ beta.exposure, ss1)
abline(a=0, b=1)
abline(a=0, b=mres1$b, col="red")

# Organise the data
dat <- tibble(
	y1 = ss1$beta.outcome,
	y2 = ss2$beta.outcome,
	x1 = ss1$beta.exposure,
	x2 = ss2$beta.exposure
)

dat$xb <- rowMeans(cbind(dat$x1, dat$x2))
dat$xb[1:100] <- 0
dat$x1[1:50] <- 0
dat$x1[101:200] <- 0
dat$x2[51:100] <- 0
dat$x2[101:200] <- 0





model <- '
y1 ~ b_1*x1 + b_1*xb + p_1*x2
y2 ~ p_2*x1 + b_2*xb + b_2*x2
x1 ~~ 0 * x2
x1 ~~ 0 * xb
x2 ~~ 0 * xb
'

mod <- sem(model, data=dat)
summary(mod)














effx <- choose_effects(50, 0.2)
effp <- (rnorm(50, sd=sd(effx)/2) + effx) %>% scale() %>% {. * sd(effx)}
cor(effx, effp)

g <- make_geno(10000, 50, 0.5)
colnames(g) <- paste0("g", 1:50)
p <- make_phen(effp, g)
x <- make_phen(effx, g)
y <- make_phen(c(0.2,-0.2), cbind(x, p))

d <- get_effs(x,y,g)
(r <- mr(d, method="mr_ivw"))
mr_scatter_plot(r, d)

library(glue)
mod <- glue('
	x ~ {paste(paste0("g", 1:50), collapse="+")}
	y ~ b * x
')
dat <- data.frame(g, x=x, y=y)
m <- sem(mod, data=dat)
summary(m)

summary(lm(y ~ x))














n <- 100
d <- tibble(
	g1 = rnorm(n),
	g2 = rnorm(n),
	gb = rnorm(n),
	x1 = rnorm(n),
	x2 = rnorm(n),
	xb = rnorm(n),
	y1 = rnorm(n),
	y2 = rnorm(n)
)

m <- '
y1 ~ b1 * x1 + b1 * xb
y2 ~ b2 * x2 + b2 * xb
x1 ~ g1
xb ~ gb
x2 ~ g2
g1 ~~ 0 * gb
g1 ~~ 0 * g2
g2 ~~ 0 * gb
'

mod <- sem(m, data=d)
semPlot::semPaths(mod, layout="tree2")
semPlot::semPaths(mod, layout="circle")


lay <- tidySEM::get_layout(
"g1", "", "gb", "", "g2",
"x1", "", "xb", "", "x2",
"", "y1", "", "y2", "", rows=3
)

tidySEM::graph_sem(mod, layout=lay)

```





```