---
title: Evaluating expected replication rates
author: Gibran Hemani
date: 2023-02-21
---


Magosi et al 2017 define a method called the M-statistic which tests if a study has systematic heterogeneity in a meta analysis, by finding the mean deviation of effects across a set of loci. Implemented in the `getspres` R package.

```{r}
library(getspres)
heartgenes214
```

```{r}
heartgenes3 <- subset(heartgenes214, 
    variants %in% c("rs10139550", "rs10168194", "rs11191416")) 
heartgenes3
```


```{r}
getspres_results <- getspres(beta_in = heartgenes3$beta_flipped, 
                               se_in = heartgenes3$gcse, 
                      study_names_in = heartgenes3$studies, 
                    variant_names_in = heartgenes3$variants)
str(getspres_results)
```

```{r}
df_spres <- getspres_results$spre_dataset
head(df_spres)
plotspres_res <- plotspres(beta_in = df_spres$beta, 
                              se_in = df_spres$se, 
                     study_names_in = as.character(df_spres$study_names), 
                   variant_names_in = as.character(df_spres$variant_names),
                           spres_in = df_spres$spre,
                          save_plot = FALSE)

```

1. Get ld region
2. simulate genetic effect in EUR
3. no genetic effect in EAS
4. meta-analysis to choose best SNP
5. raw to choose EUR SNP
6. Compare replication rate


```{r}
sim <- function(ldmat1, ldmat2, beta1)
{
  
}

```

















