---
title: "SBP scaling"
output: html_notebook
---

Systolic blood pressure has two GWASs in EAS from BBJ. How comparable are these units, and also compared against UKBB European results

```{r}
library(TwoSampleMR)
library(dplyr)
library(ieugwasr)
library(readr)
library(ggplot2)
```

Read in list of cross-ancestry SNPs to use for SBP

```{r}
snps <- read.csv("~/Downloads/bbj_new_sumstat.csv")
head(snps)
```

These are the two BBJ SBP GWASs

```{r}
gwasinfo(c("ieu-b-5075", "bbj-a-52")) %>% glimpse()
```

Relationship between these two BBJ datasets

```{r}
a <- extract_outcome_data(snps$rsid, "ieu-b-5075") %>% convert_outcome_to_exposure()
b <- extract_outcome_data(snps$rsid, "bbj-a-52")
dat <- harmonise_data(a,b)
r <- mr(dat)
r
mr_scatter_plot(r, dat)
```

The simple mode gives a very good indication here that the units are the same in the two studies. However, there is a massive outlier introduced in the more recent one, which distorts the results for the other datasets

Relationship between BBJ and UKBB-EUR

```{r}
c <- extract_outcome_data(snps$rsid, "ukb-b-20175")
dat2 <- harmonise_data(a, c)
res2 <- mr(dat2)
res2
mr_scatter_plot(res2, dat2)
```
Similar situation here where the simple mode gets it right, and the estimate suggests the scales are the same. But these massive outliers pose a problem

```{r}
c <- extract_outcome_data(snps$rsid, "ukb-b-20175")
dat3 <- harmonise_data(b%>% convert_outcome_to_exposure(), c)
res3 <- mr(dat3)
res3
mr_scatter_plot(res3, dat3)
```

Perform MR of SBP

```{r}
a <- extract_outcome_data(snps$rsid, c("ukb-b-20175", "bbj-a-52", "ieu-b-5075")) %>% convert_outcome_to_exposure()
b <- extract_outcome_data(snps$rsid, c("ebi-a-GCST005195", "bbj-a-159"))
dat <- harmonise_data(a, b)

res <- mr(dat, method_list=c("mr_ivw"))
ggplot(res, aes(x=b, y=id.outcome)) +
  geom_errorbarh(aes(xmin=b - 1.96*se, xmax=b+1.96*se), height=0) +
  geom_point(aes(colour=id.exposure)) +
  facet_grid(id.exposure ~ .) + scale_colour_brewer(type="qual")
```


```{r}
gendat <- function(snplist, x, y)
{
  a <- extract_outcome_data(snplist, x) %>% convert_outcome_to_exposure()
  b <- extract_outcome_data(snplist, y)
  dat <- harmonise_data(a, b)
  return(dat)
}

d <- gendat(snps$rsid, "ieu-b-5075", "ebi-a-GCST005195")
dr <- d %>% mr()
dr
mr_scatter_plot(dr, d)
```

```{r}
d <- gendat(snps$rsid, "ieu-b-5075", "bbj-a-159")
dr <- d %>% mr()
dr
mr_scatter_plot(dr, d)

```

That outlier SNP is just not present in BBJ's CAD.

Check concordance of outcome datasets (note that don't have SNPs selected for multiple ancestries here)



```{r}
do1 <- make_dat("ebi-a-GCST005195", "bbj-a-159")
r1 <- mr(do1)
r1
mr_scatter_plot(r1, do1)
```

```{r}
do2 <- make_dat("bbj-a-159", "ebi-a-GCST005195")
r2 <- mr(do2)
r2
mr_scatter_plot(r2, do2)
```

Using European instruments, there is a large effect in EUR that doesn't replicate in EAS. However the simple mode ignors this and gives a slope of ~1. Using EAS instruments gives two apparent clusters, one with large effects in EAS that are null in EAS. Those are likely due to LD. A visual inspection of the other cluster indicates a slope of ~1. I would not adjust this dataset.

**Use the better powered GWAS as the exposure to reduce regression dilution bias**

```{r}
adjx <- gendat(snps$rsid, "ukb-b-20175", "bbj-a-52") 
r <- mr(adjx)
r
mr_scatter_plot(r, adjx)
```


The effect appears to be quite convincingly attenuated by ~20% in BBJ







